---
title: "Stat 433 Project"
author: "Cameron Jones"
date: "11/16/2021"
output: html_document
---

```{r setup, include=FALSE, results = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(lmerTest))
```

# Scroll to bottom to see figure 1!

## Thesis

Among Premier League teams over the last seven years, the club transfer windows with higher net transfer spend have higher increase in expected points from season to season compared to club transfer windows with lower net transfer spend, controlling for average age of transfer.


## Reading Data Into R

### Performance Data

season stats, goals, expected goals, etc., from understat:
https://www.kaggle.com/slehkyi/extended-football-stats-for-european-leagues-xg

```{r, message = FALSE}
performanceData = read_csv("ProjectData\\understat.csv")
str(performanceData)
```

### Transfer Data

Individual transfers in European soccer leagues as scraped from www.transfermarkt.com, by github user ewenme: https://github.com/ewenme/transfers

```{r, message = FALSE}
rawTransferData = read_csv("https://raw.githubusercontent.com/ewenme/transfers/master/data/2014/english_premier_league.csv")

for (i in 2015:2019) {
  link = paste0("https://raw.githubusercontent.com/ewenme/transfers/master/data/", i, "/english_premier_league.csv")
  data = read_csv(link)
  rawTransferData = bind_rows(rawTransferData, data)
}
str(rawTransferData)
```

## Data Cleaning and Preparing

### Transfer Data

```{r, message = FALSE}
transferData = rawTransferData

#adjust data formatting issues
transferData$fee_cleaned[transferData$player_name == "Bertrand Traoré" & transferData$age == 20] = 1.8
transferData$club_name = sub(" FC", "", transferData$club_name)
transferData$club_name[transferData$club_name == "AFC Bournemouth"] = "Bournemouth"
transferData$club_name[transferData$club_name == "Brighton & Hove Albion"] = "Brighton"
transferData$club_name[transferData$club_name == "Cardiff City"] = "Cardiff"
transferData$club_name[transferData$club_name == "Huddersfield Town"] = "Huddersfield"
transferData$club_name[transferData$club_name == "Hull City"] = "Hull"
transferData$club_name[transferData$club_name == "Leicester City"] = "Leicester"
transferData$club_name[transferData$club_name == "Norwich City"] = "Norwich"
transferData$club_name[transferData$club_name == "Stoke City"] = "Stoke"
transferData$club_name[transferData$club_name == "Sunderland AFC"] = "Sunderland"
transferData$club_name[transferData$club_name == "Swansea City"] = "Swansea"
transferData$club_name[transferData$club_name == "Tottenham Hotspur"] = "Tottenham"
transferData$club_name[transferData$club_name == "West Ham United"] = "West Ham"

#filter to summer transfers with fees; control for transfer in or out
transferData$fee_cleaned[transferData$transfer_movement == "out"] = -1 * transferData$fee_cleaned
transferData = transferData %>% filter(transfer_period == "Summer", complete.cases(fee_cleaned))

#Classify by three groups: Attacker, Midfielder, Defender
transferData$positionGroup = "Attacker"
transferData$positionGroup[grepl("Midfield", transferData$position)] = "Midfielder"
transferData$positionGroup[(grepl("Back", transferData$position) | transferData$position == "Goalkeeper")] = "Defender"

#dplyr 
transferData = transferData %>% group_by(year, club_name, positionGroup) %>% summarize(netTransferSpend = sum(fee_cleaned), avgAge = mean(age), n = n()) %>% pivot_wider(names_from = positionGroup, values_from = c(netTransferSpend, avgAge, n))

transferData[is.na(transferData)] = 0 #some teams did not buy or sell a certain class of player in a given summer. set these to zero.

transferData = transferData %>% rename(team = club_name) %>% mutate(netTransferSpend_Total = sum(netTransferSpend_Attacker, netTransferSpend_Midfielder, netTransferSpend_Defender))

```


### Performance Data

```{r, message = FALSE}
#Selects the Premier League teams and drops the "league" column.
performanceData = performanceData %>% filter(...1 == "EPL") %>% rename(year = ...2) %>% select(-...1)
pd = performanceData
pd["lastYearXpts"] = 0
pd["lastYearpts"] = 0

#Adding last year's points to this year's row

for (i in 2015:2019) {
  listofTeams = pd[pd["year"] == i, "team"]
  for (j in 1:nrow(listofTeams)) {
    t = toString(listofTeams[j,])
    #cat("year: ", i, " team: ", t, "\n\n")
    lastYearXP = as.numeric(pd[(pd["year"] == i-1 & pd["team"] == t), "xpts"])
    pd[(pd["year"] == i & pd["team"] == t), "lastYearXpts"] = lastYearXP
  }
}

# Take out promoted teams and 2014 data, because we have no 2013 data
pd = pd[which(!(pd$lastYearXpts == 0) & !is.na(pd$lastYearXpts)),]
```

## Results

```{r, message = FALSE}
finalData = left_join(pd, transferData) %>% mutate(avgAge_Overall = (avgAge_Attacker * n_Attacker + avgAge_Midfielder * n_Midfielder + avgAge_Defender * n_Defender)/(n_Attacker + n_Midfielder + n_Defender), diffXP = xpts - lastYearXpts) %>% select(year, team, pts, xpts, lastYearXpts, diffXP, xG, xGA, contains(c("net", "avg", "n_")))

finalData$team = as.factor(finalData$team)

#summary(lm(diffXP ~ netTransferSpend_Total + avgAge_Overall, data = finalData))
fit = lmer(diffXP ~ netTransferSpend_Total + avgAge_Overall + (1 | team), data = finalData)
summary(fit)
rand(fit)
```

### Figure 1

```{r}
finalData %>% ggplot(aes(x = netTransferSpend_Total, y = diffXP)) +
  geom_point() +
  geom_smooth(method = "lm", se = T, size = 3) +
  ggtitle("Increase in xP by Summer Net Transfer Spend, 2015-2019") + 
  xlab("Net Transfer Spend (Millions, £)") +
  ylab("Increase in xP")
```

